Blame and Coercion: Together Again for the Second Time

### 2019-10-09 paper.tex
hyper-coercion version of applyCast

\newtheorem{theorem}{Theorem}[]
\newtheorem{proposition}{Proposition}[]
\newtheorem{corollary}{Corollary}[]
\newcommand{\figref}[1]{Fig.~\ref{#1}}
\newcommand{\stxrule}[3]{#1 & ::= & #3 & \text{#2}\\}
\newcommand{\stxrulecont}[1]{& | & #1 & \\}
\newcommand{\funrule}[3]{#1 &=& #2 & #3\\}
\newcommand{\redrule}[3]{#1 & \longrightarrow_B & #2 & #3\\}
\newcommand{\comprule}[4]{#1 & \fatsemi^\ell & #2 & = & #3 & #4 \\}
\newcommand{\plus}[0]{+}
\newcommand{\judgetype}[3]{#1 \vdash #2 : #3}
\newcommand{\lazyUD}{$\mathtt{Lazy UD}$}
\newcommand{\lazyD}{$\mathtt{Lazy D}$}
\newcommand{\eager}{$\mathtt{Eager}$}
\newcommand{\hyperCoercionI}[0]{id\star}
\newcommand{\hyperCoercionC}[3]{#1 \overset{#2}{\curvearrowright} #3}
\newcommand{\sOOinspect}[3]{\mathtt{Eval} \; #1 \; #2 \; #3}
\newcommand{\sOOreturn}[2]{\mathtt{Apply} \; #2 \; #1}
\newcommand{\sOOhalt}[1]{\mathtt{Halt} \; #1}
\newcommand{\TOOdyn}[0]{\star}
\newcommand{\TOOpre}[1]{#1}
\newcommand{\POOunit}[0]{\iota}
\newcommand{\POOfun}[2]{#1 \shortrightarrow #2}
\newcommand{\POOprod}[2]{#1 \times #2}
\newcommand{\POOsum}[2]{#1 \plus #2}
\newcommand{\eOOvar}[1]{#1}
\newcommand{\eOOsole}[0]{\mathtt{tt}}
\newcommand{\eOOlam}[4]{\lambda^{#1\rightarrow{}#2}#3.#4}
\newcommand{\eOOapp}[2]{#1 \; #2}
\newcommand{\eOOcons}[2]{\mathtt{cons} \; #1 \; #2}
\newcommand{\eOOcar}[1]{\mathtt{car} \; #1}
\newcommand{\eOOcdr}[1]{\mathtt{cdr} \; #1}
\newcommand{\eOOinl}[1]{\mathtt{inl} \; #1}
\newcommand{\eOOinr}[1]{\mathtt{inr} \; #1}
\newcommand{\eOOcase}[3]{\mathtt{case} \; #1 \; #2 \; #3}
\newcommand{\eOOcast}[4]{#1 \langle \cOOcast{#2}{#3}{#4} \rangle}
\newcommand{\eOOblame}[1]{\mathtt{blame} \; #1}
\newcommand{\oOOinj}{\mathtt{dyn}}
\newcommand{\oOOsole}{\mathtt{tt}}
\newcommand{\oOOfun}{\mathtt{fun}}
\newcommand{\oOOcons}{\mathtt{cons}}
\newcommand{\oOOinl}{\mathtt{inl}}
\newcommand{\oOOinr}{\mathtt{inr}}
\newcommand{\oOOblame}[1]{\mathtt{blame} \; #1}
\newcommand{\cOOcast}[3]{#1 \overset{#2}{\Rightarrow} #3}
\newcommand{\vOOcast}[2]{#1\langle#2\rangle}
\newcommand{\vOOfun}[3]{\mathtt{fun} \; #1 \; #2 \; #3}
\newcommand{\vOOtt}[0]{\mathtt{tt}}
\newcommand{\vOOcons}[2]{\mathtt{cons}\;#1\;#2}
\newcommand{\vOOinl}[1]{\mathtt{inl}\;#1}
\newcommand{\vOOinr}[1]{\mathtt{inr}\;#1}
\newcommand{\rOOsucc}[1]{\mathtt{succ}\;#1}
\newcommand{\rOOfail}[1]{\mathtt{fail}\;#1}

@@

subtyping, reduction ...

@@

\begin{figure}
	\[ 
	\begin{array}{lclr}
	\stxrule{c}{hyper-coercions}{
		\hyperCoercionI \mid{}
		\hyperCoercionC{h}{m}{t}
	}
	\stxrule{h}{heads}{
		\epsilon \mid{}
		?^l
	}
	\stxrule{m}{middles}{
		U \mid{}
		c_1 \rightarrow c_2 \mid{}
		c_1 \times c_2 \mid{}
		c_1 \plus c_2
	}
	\stxrule{t}{tails}{
		\epsilon \mid{}
		! \mid{}
		\bot^l
	}
	\stxrule{\ell}{$ \mathtt{Maybe} \; l $}{
		\epsilon \mid{}
		l
	}
	\end{array}
	\]

	\fbox{$ c \fatsemi^\ell c = c $}
	\[ 
	\begin{array}{rclclr}

@@

	\fbox{$ seq(c,c) = c $}
	\[
	\begin{array}{rclr}
	\funrule{seq(c_1,c_2)}{
		c_1 \fatsemi^\epsilon c_2
	}{cod(c_1) = dom(c_2)}
	\end{array}
	\]

@@

	\fbox{$ cast(T,l,T) = c$}
	\[
	\begin{array}{rclr}
	\funrule{cast(T_1,l,T_2)}{
		id(T_1) \fatsemi^l id(T_2)
	}{}
	\end{array}
	\]

	\caption{$Lazy D$ Hyper-coercion}
	\label{fig:HC-D}
\end{figure}

\subsection{$Lazy D$ Hyper-coercion}

The syntax of \lazyD{} Hyper-coercion is shown in \figref{fig:HC-D}.

>>>>

\newtheorem{theorem}{Theorem}[]
\newtheorem{proposition}{Proposition}[]
\newtheorem{corollary}{Corollary}[]
\newcommand{\figref}[1]{Fig.~\ref{#1}}
\newcommand{\stxrule}[3]{#1 & ::= & #3 & \text{#2}\\}
\newcommand{\stxrulecont}[1]{& | & #1 & \\}
\newcommand{\funrule}[3]{#1 &=& #2 & #3\\}
\newcommand{\redrule}[3]{#1 & \longrightarrow_B & #2 & #3\\}
\newcommand{\comprule}[4]{#1 & \fatsemi^\ell & #2 & = & #3 & #4 \\}
\newcommand{\plus}[0]{+}
\newcommand{\judgetype}[3]{#1 \vdash #2 : #3}
\newcommand{\lazyUD}{$\mathtt{Lazy UD}$}
\newcommand{\lazyD}{$\mathtt{Lazy D}$}
\newcommand{\eager}{$\mathtt{Eager}$}
\newcommand{\hyperCoercionI}[0]{\mathtt{id\star}}
\newcommand{\hyperCoercionC}[3]{#1 \overset{#2}{\curvearrowright} #3}
\newcommand{\sOOinspect}[3]{\mathtt{Eval} \; #1 \; #2 \; #3}
\newcommand{\sOOreturn}[2]{\mathtt{Apply} \; #2 \; #1}
\newcommand{\sOOhalt}[1]{\mathtt{Halt} \; #1}
\newcommand{\TOOdyn}[0]{\star}
\newcommand{\TOOpre}[1]{#1}
\newcommand{\POOunit}[0]{\iota}
\newcommand{\POOfun}[2]{#1 \shortrightarrow #2}
\newcommand{\POOprod}[2]{#1 \times #2}
\newcommand{\POOsum}[2]{#1 \plus #2}
\newcommand{\eOOvar}[1]{#1}
\newcommand{\eOOsole}[0]{\mathtt{tt}}
\newcommand{\eOOlam}[4]{\lambda^{#1\rightarrow{}#2}#3.#4}
\newcommand{\eOOapp}[2]{#1 \; #2}
\newcommand{\eOOcons}[2]{\mathtt{cons} \; #1 \; #2}
\newcommand{\eOOcar}[1]{\mathtt{car} \; #1}
\newcommand{\eOOcdr}[1]{\mathtt{cdr} \; #1}
\newcommand{\eOOinl}[1]{\mathtt{inl} \; #1}
\newcommand{\eOOinr}[1]{\mathtt{inr} \; #1}
\newcommand{\eOOcase}[3]{\mathtt{case} \; #1 \; #2 \; #3}
\newcommand{\eOOcast}[4]{#1 \langle \cOOcast{#2}{#3}{#4} \rangle}
\newcommand{\eOOblame}[1]{\mathtt{blame} \; #1}
\newcommand{\oOOinj}{\mathtt{dyn}}
\newcommand{\oOOsole}{\mathtt{tt}}
\newcommand{\oOOfun}{\mathtt{fun}}
\newcommand{\oOOcons}{\mathtt{cons}}
\newcommand{\oOOinl}{\mathtt{inl}}
\newcommand{\oOOinr}{\mathtt{inr}}
\newcommand{\oOOblame}[1]{\mathtt{blame} \; #1}
\newcommand{\cOOcast}[3]{#1 \overset{#2}{\Rightarrow} #3}
\newcommand{\vOOcast}[2]{#1\langle#2\rangle}
\newcommand{\vOOfun}[3]{\mathtt{fun} \; #1 \; #2 \; #3}
\newcommand{\vOOtt}[0]{\mathtt{tt}}
\newcommand{\vOOcons}[2]{\mathtt{cons}\;#1\;#2}
\newcommand{\vOOinl}[1]{\mathtt{inl}\;#1}
\newcommand{\vOOinr}[1]{\mathtt{inr}\;#1}
\newcommand{\rOOsucc}[1]{\mathtt{succ}\;#1}
\newcommand{\rOOfail}[1]{\mathtt{fail}\;#1}
\newcommand{\typingHC}[3]{#1 : #2 \Longrightarrow #3}
\newcommand{\hcvOOinj}[2]{\mathtt{inj} \; #1 \; #2}
\newcommand{\hcvOOfun}[5]{\mathtt{fun} \; #1 \; #2 \; #3 \; #4 \; #5}
\newcommand{\hcvOOtt}[0]{\mathtt{tt}}
\newcommand{\hcvOOcons}[4]{\mathtt{cons}\;#1\;#2\;#3\;#4}
\newcommand{\hcvOOinl}[2]{\mathtt{inl}\;#1\;#2}
\newcommand{\hcvOOinr}[2]{\mathtt{inr}\;#1\;#2}

@@

\dots

@@

\begin{figure}
	\[ 
	\begin{array}{lclr}
	\stxrule{c}{hyper-coercions}{
		\hyperCoercionI \mid{}
		\hyperCoercionC{h}{m}{t}
	}
	\stxrule{h}{heads}{
		\epsilon \mid{}
		?^l
	}
	\stxrule{m}{middles}{
		\POOunit \mid
		\POOfun{c_1}{c_2} \mid
		\POOprod{c_1}{c_2} \mid
		\POOsum{c_1}{c_2}
	}
	\stxrule{t}{tails}{
		\epsilon \mid{}
		! \mid{}
		\bot^l
	}
	\stxrule{\ell}{$ \mathtt{Maybe} \; l $}{
		\epsilon \mid{}
		l
	}
	\end{array}
	\]

	Coercion typing \fbox{$ c : T \Longrightarrow T $}
	\begin{gather*}
	\inference{}{\typingHC{\hyperCoercionI}{\TOOdyn}{\TOOdyn}}
	\quad
	\inference{
		\typingHC{h}{T_1}{P_1} &
		\typingHC{m}{P_1}{P_2} &
		\typingHC{t}{P_2}{T_2}
	}{
		\typingHC{\hyperCoercionC{h}{m}{t}}{T_1}{T_2}
	}
	\end{gather*}

	Head typing \fbox{$ \typingHC{h}{T}{P} $}
	\begin{gather*}
	\inference{}{\typingHC{\epsilon}{P}{P}}
	\quad
	\inference{}{\typingHC{?^l}{\TOOdyn}{P}}
	\end{gather*}

	Middle typing \fbox{$ \typingHC{m}{P}{P} $}
	\begin{gather*}
	\inference{}{\typingHC{\POOunit}{\POOunit}{\POOunit}}
	\\
	\inference{
		\typingHC{c_1}{T_3}{T_1} &
		\typingHC{c_2}{T_2}{T_4}
	}{
		\typingHC{\POOfun{c_1}{c_2}}{\POOfun{T_1}{T_2}}{\POOfun{T_3}{T_4}}
	}
	\\
	\inference{
		\typingHC{c_1}{T_1}{T_3} &
		\typingHC{c_2}{T_2}{T_4}
	}{
		\typingHC{\POOprod{c_1}{c_2}}{\POOprod{T_1}{T_2}}{\POOprod{T_3}{T_4}}
	}
	\\
	\inference{
		\typingHC{c_1}{T_1}{T_3} &
		\typingHC{c_2}{T_2}{T_4}
	}{
		\typingHC{\POOsum{c_1}{c_2}}{\POOsum{T_1}{T_2}}{\POOsum{T_3}{T_4}}
	}
	\end{gather*}

	Tail typing \fbox{$ \typingHC{t}{P}{T} $}
	\begin{gather*}
	\inference{}{\typingHC{\epsilon}{P}{P}} \quad
	\inference{}{\typingHC{!}{P}{\TOOdyn}} \quad
	\inference{}{\typingHC{\bot^l}{P}{T}} \quad
	\end{gather*}

	\caption{Syntax of hyper-coercion}
	\label{fig:hyper-coercion}
\end{figure}

\figref{fig:hyper-coercion} defines the syntax of hyper-coercions. All 
definitions and results in this section are new.

Terms, types, pre-types, blame labels, and observations are as before. Let $ c 
$ ranges over hyper-coercions. A hyper-coercion is either an identity 
cast between the dynamic types, or a complex including a head, a middle, and a 
tail. Let $ h $ ranges over heads. A head is either a no-op, or a projection 
decorated with a label, to which blame is allocated if the projection fails. 
Let $ m $ ranges over middles. There is a one-to-one 
correspondence between middles and type constructors. Let $ t $ ranges over 
tails. A tail is either a no-op, an injection, or a failure. Let $ \ell $ 
ranges over no-op and labels. It is useful when we introduce composition of 
hyper-coercions. Typing rules for hyper-coercions, heads, middles, and tails 
are straightforward.

\subsection{\lazyD{} Hyper-coercion}

\begin{figure}
	\fbox{$ c \fatsemi^\ell c = c $}
	\[ 
	\begin{array}{rclclr}

@@

	\fbox{$ seq(c,c) = c $}
	\[
	\begin{array}{rclr}
	\funrule{seq(c_1,c_2)}{
		c_1 \fatsemi^\epsilon c_2
	}{
		c_1 : T_1 \Longrightarrow T_2
		\; \text{and} \;
		c_2 : T_2 \Longrightarrow T_3
	}
	\end{array}
	\]

@@

	\fbox{$ cast(T,l,T) = c$}
	\[
	\begin{array}{rclr}
	\funrule{cast(T_1,l,T_2)}{
		id(T_1) \fatsemi^l id(T_2)
	}{}
	\end{array}
	\]
	\caption{$Lazy D$ Hyper-coercion}
	\label{fig:HC-D}
\end{figure}

Functions that construct \lazyD{} Hyper-coercion are shown in \figref{fig:HC-D}.
When a language implementation want to construct a coercion, it should use 
these functions: $ cast(T_1,l,T_2) $, $ id(T) $, and $ seq(c_1,c_2) $. $ 
cast(T_1,l,T_2) $ constructs a coercion from a source type, a label, and a 
target type. $ id(T) $ constructs an identity cast of $ T $. $ seq(c_1,c_2) $ 
composes two coercions where the target type of $ c_1 $ is equal to the source 
type of $ T_2 $.

\begin{figure}
	\fbox{$ applyCast(c,v) = r $}
	\[
	\begin{array}{rclr}
	\funrule{applyCast(\hyperCoercionI,\;v)}{\rOOsucc{v}}{}
	\funrule{applyCast(\hyperCoercionC{?^l}{m}{t},\;\hcvOOinj{P}{v})}{
		applyTail(t',applyMiddle(m',v))
	}{id(P)\fatsemi^l(m,t) = (m',t')}
	\funrule{applyCast(\hyperCoercionC{\epsilon}{m}{t},\;v)}{
		applyTail(t,applyMiddle(m,v))}
	{}
	\end{array}
	\]

	\fbox{$ applyMiddle(m,v) = v $}
	\[
	\begin{array}{rclr}
	\funrule{applyMiddle(\POOunit,\hcvOOtt)}{\hcvOOtt}{}
	\funrule{applyMiddle(\POOfun{c_3}{c_4},\hcvOOfun{c_1}{\rho}{x}{e}{c_2})}{
		\hcvOOfun{seq(c_3,c_1)}{\rho}{x}{e}{seq(c_2,c_4)}
	}{}
	\funrule{applyMiddle(\POOprod{c_3}{c_4},\hcvOOcons{v_1}{c_1}{v_2}{c_2})}{
		\hcvOOcons{v_1}{seq(c_1,c_3)}{v_2}{seq(c_2,c_4)}
	}{}
	\funrule{applyMiddle(\POOsum{c_3}{c_4},\hcvOOinl{v}{c_1})}{
		\hcvOOinl{v}{seq(c_1,c_3)}
	}{}
	\funrule{applyMiddle(\POOsum{c_3}{c_4},\hcvOOinr{v}{c_2})}{
		\hcvOOinr{v}{seq(c_2,c_4)}
	}{}
	\end{array}
	\]

	\fbox{$ applyTail(t,v) = r $}
	\[
	\begin{array}{rclr}
	\funrule{applyTail(\epsilon,v)}{\rOOsucc{v}}{}
	\funrule{applyTail(!,v)}{\rOOsucc{(\hcvOOinj{P}{v})}}{v : P}
	\funrule{applyTail(\bot^l,v)}{\rOOfail{l}}{}
	\end{array}
	\]
	\caption{Applying hyper-coercions to values}
\end{figure}

\begin{theorem}[$ id $ constructs neutrals] For all $ c : T_1 \Longrightarrow 
T_2 $
	\begin{enumerate}
		\item $ seq(id(T_1),c) = c $
		\item $ seq(c,id(T_2)) = c $
	\end{enumerate}
\end{theorem}
